What's in a Name? \\ Linear Temporal Logic \\ Literally Represents Time Lines

### 2023-08-27 paper/main.tex
more typo fixes...

\begin{definition}[Linear Temporal Logic (\ltl)]
    % cite Kristin's survey paper
    The syntax of an \ltl\ formula over a set of atomic propositions $\AP$, where $p\in\AP$ is a propositional variable, consists of the following grammar:
    \begin{align*}
        \varphi ::= & p \mid \neg \varphi \mid \varphi \land \psi \mid \varphi \lor \psi \mid \varphi \limplies \psi \mid \always \varphi \mid \eventually \varphi \mid \nextt \varphi \\ & \mid \varphi \stronguntil \psi \mid \varphi \weakrelease \psi
    \end{align*}\label{ltl-defn}
\end{definition}

@@

\subsection{State-based \Buchi automata (BA)}
\begin{definition}[$\omega$-word]
    Let $\Sigma$ be an alphabet. An $\omega$-word or infinite run of $\Sigma$, is an infinite string $s = (s_0, s_1, s_2, \dots)$ where each $s_i\in \Sigma$.
\end{definition}
\begin{definition}[\Buchi automaton]
    A \Buchi automaton is a $5$-tuple, $B = (Q, \Sigma, \delta, s, F)$ consisting of
    \begin{enumerate}
        \item a finite set of states $Q$
        \item a finite alphabet of input symbols $\Sigma$
        \item a transition relation $\delta \subseteq (Q\times \Sigma) \times Q$
        \item an initial or start state called $s\in Q$
        \item a set of accepting states $F \subseteq Q$
    \end{enumerate}
    $B$ accepts an infinite run iff at least one of its infinitely visited states is in $F$.
\end{definition}

\subsection{$\omega$-regular expressions}
\begin{definition}[Regular expression]\label{def:regex}
    Let $a$ be a symbol in $\Sigma$, $A$ be a regular expression, and $\epsilon$ be the empty expression. We define regular expressions using the following grammar:
    \begin{align*}
        A ::= \emptyset \mid \epsilon \mid a \mid AA \mid A + A \mid A^*
    \end{align*}
\end{definition}
\begin{definition}[Semantics of regular expressions]
    Let $\L(A)$ denote the set of propositional formulas accepted by regular expression $A$, also called the language accepted by $A$. We define $\L(A)$, inductively as:
    \begin{align*}
        \L(\emptyset) & = \emptyset \\
        \L(\epsilon) & = \{\epsilon\} \tag{$\epsilon$ denotes the empty string}\\
        \L(a) & = \{a\} \\
        \L(A_1A_2) & = \{s_1s_2 \mid s_1 \in L(A_1) \text{ and } s_2 \in L(A_2)\} \\
        \L(A_1 + A_2) & = L(A_1) \cup L(A_2) \\
        \L^{(0)}(A) & = \{\epsilon\} \\
        \L^{(i + 1)}(A) & = \{s_1s_2 \mid s_1 \in L(A) \text{ and } s_2 \in L^{(i)}(A)\} \\\
        \L(A^*) & = \bigcup_{i \ge 0} L^{(i)}(A)
    \end{align*}
\end{definition}

\begin{remark}
    For the purpose of our tool, we define $\Sigma$ as the set of propositional logic formulas:
    \[
    a \in \Sigma ::= p \mid \top \mid \bot \mid \neg \phi \mid \phi \land \psi \mid \phi \lor \psi \mid \phi \limplies \psi
    \]
\end{remark}

\begin{definition}[$\omega$-regular expression]\label{def:omega-regex}
    Regular expression concerns only finite-length strings. However, since \ltl\ formulas reason about events that happen over an infinite-length timeline, we need to model it using infinite regular expressions (i.e., $\omega$-regular expression). $\omega$-regular expression is defined by the following grammar:
    \begin{align*}
        B ::= A^{\omega} \mid AB \mid B + B
    \end{align*}
\end{definition}
\begin{definition}[Semantics of $\omega$-regular expression]\label{def:omega-semantics}
    Let $\Sigma^{\omega}$ denote the set of infinite-length strings over fixed alphabet $\Sigma$. Let $L_{\omega}(B)$ denote the $\omega$-language accepted by $\omega$-regular expression $B$. Then $L_{\omega}(B)$ is inductively defined as
    \begin{align*}
        L_{\omega}(A^{\omega}) & = \{s_1s_2s_3\cdots \mid s_i \in L(A) \text{ and } i \ge 1\} \tag{$\epsilon
         \not\in L(A)$} \\
        L_{\omega}(AB) & = \{s_1s_2 \mid s_1 \in L(A) \text{ and } s_2 \in L_{\omega}(B)\} \\
        L_{\omega}(B_1 + B_2) &= L_{\omega}(B_1) \cup L_{\omega}(B_2)
    \end{align*}
\end{definition}

@@

\subsection{$\omega$-regex to timeline} \label{regex2timeline}
Our tool uses Graphviz~\cite{Ellson2001GraphvizO} to achieve the timeline visualization step. By construction of our algorithm, every $\omega$-regular expression is in the form of
\[
    A_1A_2^{\omega} + A_3A_4^{\omega} + \cdots + A_{2n-1}A_{2n}^{\omega}
\]
where $A_i$ are regular expressions, $A_{2i-1}$ could be $\epsilon$, and $\epsilon \not\in L(A_{2i})$. At a high level, for each of regular expression $A_i$, we visualize it as a set of accepted input. We view each union operator as parallel timelines, and each $A_{2i-1}$ is concatenated with $A_{2i}$ repeated infinitely many times, as represented in the grey box. \cref{fig:timeline} presents a generic timeline resulting from this form of construction.
\begin{figure}[h!]
    \centering
    \begin{tikzpicture}
    \node[draw, thick, fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a1) {$A_1$};
    \node[shape=rectangle, fill=gray!30!white, style={minimum width=1.8cm, minimum height=1.3cm}] (a2g) [right= 0.85cm of a1] {};
    \node[fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a2) [right= 1cm of a1] {$A_2$};
    \node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a3) [below= 0.4cm of a1] {$A_3$};
    \node[shape=rectangle, fill=gray!30!white, style={minimum width=1.8cm, minimum height=1.3cm}] (a4g) [right= 0.85cm of a3] {};
    \node[fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a4) [right= 1cm of a3] {$A_4$};
    \node[draw, thick, fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a5) [below= 0.4cm of a3] {...};
    \node[draw, thick, fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a7) [below= 0.4cm of a5] {\!\!\!$A_{2n-1}$\!\!\!};
    \node[shape=rectangle, fill=gray!30!white, style={minimum width=1.8cm, minimum height=1.3cm}] (a8g) [right= 0.85cm of a7] {};
    \node[fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a8) [right= 1cm of a7] {\!\!\!$A_{2n}$\!\!\!};
    \node (m) at ($(a1)!0.5!(a7)$) {};
    \node[draw, thick, fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (s) [left= 1.5cm of m] {\!\!\!\!start\!\!\!\!};
    \draw[thick] (s) edge[-latex] (a1) (s) edge[-latex] (a3) (s) edge[-latex] (a5) (s) edge[-latex] (a7);
    \draw[thick] (a1) edge[-latex] (a2) (a3) edge[-latex] (a4) (a7) edge[-latex] (a8);
    \end{tikzpicture}
%    \includegraphics[scale=0.5]{img/timeline.png}
    \caption{Generic timeline construction from $A_1A_2^{\omega} + A_3A_4^{\omega} + \cdots + A_{2n-1}A_{2n}^{\omega}$}
    \label{fig:timeline}
\end{figure}

@@

\begin{lemma}
    For any NFA $G$ with start state $s$ and (exactly one) final state $f$, \textbf{nfa2regex\_firstvisit}$(G, s, f)$ outputs the regular expression of all paths from $s$ to $f$, visiting $f$ for the first time.
\end{lemma}
\begin{proof}
    Let $G'$ be $G$ with all out-edge (including loops) from $f$ deleted. We claim that $L(G')$ is equal to the language $L'$ of all paths from $s$ reaching $f$ for the first time.

    First we show $L(G') \subseteq L'$, i.e., every path accepted by $G'$ starts in $s$ and ends by reaching $f$ for the first time. Since $s$ and $f$ are the initial and final states of $G'$ respectively, clearly every path accepted by $G$ must start in $s$ and end in $f$. Further, as $f$ has no out-edges in $G$, any accepted path must end immediately once it reaches $f$. Thus it must end once it reaches $f$ for the first time.

    Also, we show $L' \subseteq L(G')$, i.e., every path starting in $s$ and ending by $f$ reaching $f$ for the first time is accepted by $G'$. If a path contains an out-edge of $f$, it cannot be in $L'$ as the trace continues beyond the first time step at $f$. Thus, every trace in $L'$ only uses edges in $G'$, and is thus accepted by $G$ iff it is accepted by $G'$.
\end{proof}

\begin{theorem}
    For any \Buchi automaton $G$ with start state $s$ and final states $F$, \textbf{ba2wregex}$(G)$ outputs the $\omega$-regular language accepted by $G$
\end{theorem}
\begin{proof}
    We show that
    \[
        L = \bigcup_{f\in F} \textbf{nfa2regex}(G, s, f) \verb|^| \left(\textbf{nfa2regex\_firstvisit}(G, f, f)\right)^\omega
    \]
    and the $\omega$-language accepted by $G$, $L_\omega(G)$ are identical.

    First, $L\subseteq L_\omega(G)$ since for every infinite run in $L$, clearly by definition of $L$, there is some $f\in F$ such that passes through $f$ infinitely often. The converse is true as well: for every infinite run $q_0, q_1, \dots$ accepted by $G$, then $q_0 = s$ and by definition of the accepting language, there is some accepting state $f$, passed through infinitely often, thus the run is in $\textbf{nfa2regex}(G, s, f)$ \verb|^| $\left(\textbf{nfa2regex\_firstvisit}(G, f, f)\right)^\omega$
\end{proof}
\subsection{Correctness of Rewrite Rules}
\begin{lemma}
    For all regular expressions $r_1, r_2, r$, each of the following rewrite rules preserve the regular or $\omega$-regular language represented by the expressions:
    \begin{alignat}{2}
        & r_1 + r_1r_2^* && \Longrightarrow r_1r_2^* \\
        & r + r && \Longrightarrow r \\
        & r_1 + r_2^*r_1 && \Longrightarrow r_2^*r_1 \\
        & (r^*)^{\omega} && \Longrightarrow r^{\omega} \\
        & (r_1r_2^*)r_2^{\omega} && \Longrightarrow r_1r_2^{\omega} \\
        & (r_1r_2)r_2^{\omega} && \Longrightarrow r_1r_2^{\omega} \\
        & r^*r^{\omega} && \Longrightarrow r^{\omega} \\
        & rr^{\omega} && \Longrightarrow r^{\omega}
    \end{alignat}
\end{lemma}

@@

\appendix
\section{Concrete Syntax of \tool} \label{sec:concrete-syntax}
Recall from~\ref{ltl-defn} that we define an \ltl\ formula as
\begin{align*}
    \varphi ::= & p \mid \neg \varphi \mid \varphi \land \psi \mid \varphi \lor \psi \mid \varphi \limplies \psi \mid \always \varphi \mid \eventually \varphi \mid \nextt \varphi \\ & \mid \varphi \stronguntil \psi \mid \varphi \weakrelease \psi
\end{align*}
where $p \in \AP$ and $\varphi$ and $\psi$ are LTL formulas. In our tool \tool, we use the following concrete input syntax to represent \ltl\ formulas in ASCII text.
\[
    \begin{array}{c r l l l@{\qquad} l}
         &      & \textit{abstract syntax}              & \textit{concrete syntax}                       & \textit{description} \\
    \varphi & ::=  & p                                     & \texttt{p}                                     & \text{atomic prop.} \\
         & \mid & \neg \varphi                             & \texttt{!} \varphi                                & \text{negation} \\
         & \mid & \varphi \land \psi                       & \varphi \texttt{ \& } \psi                        & \text{conjunction} \\
         & \mid & \varphi \lor \psi                        & \varphi \texttt{ | } \psi                         & \text{disjunction} \\
         & \mid & \varphi \limplies \psi                   & \varphi \texttt{ -> } \psi                        & \text{implication} \\
         & \mid & \always \varphi                          & \texttt{G} \varphi                                & \text{globally} \\
         & \mid & \eventually \varphi                      & \texttt{F} \varphi                                & \text{in the future} \\
         & \mid & \nextt \varphi                           & \texttt{X} \varphi                                & \text{next} \\
         & \mid & \varphi \stronguntil \psi                & \varphi \texttt{ U } \psi                         & \text{until} \\
         & \mid & \varphi \weakrelease \psi                & \varphi \texttt{ R } \psi                         & \text{release} \\
    \end{array}
\]
%\nocite{*}
\bibliographystyle{IEEEtran}
\bibliography{RelatedWork}

>>>>

\begin{definition}[Linear Temporal Logic (\ltl)]
    % cite Kristin's survey paper
    The syntax of an \ltl\ formula over a set of atomic propositions $\AP$, where $p\in\AP$ is a propositional variable, consists of the following grammar:
    \begin{align*}
        \varphi = & p \mid \neg \varphi \mid \varphi \land \psi \mid \varphi \lor \psi \mid \varphi \limplies \psi \mid \always \varphi \mid \eventually \varphi \mid \nextt \varphi \\ & \mid \varphi \stronguntil \psi \mid \varphi \weakrelease \psi
    \end{align*}\label{ltl-defn}
\end{definition}

@@

\subsection{State-based \Buchi automata (BA)}
\begin{definition}[$\omega$-word]
    Let $\Sigma$ be an alphabet. An $\omega$-word or infinite run of $\Sigma$, is an infinite string $s = (s_0, s_1, s_2, \dots)$ where each $s_i\in \Sigma$.
\end{definition}
\begin{definition}[\Buchi automaton (BA)]
    A \Buchi automaton is a $5$-tuple, $(Q, \Sigma, \delta, s, F)$ consisting of
    \begin{enumerate}
        \item a finite set of states $Q$
        \item a finite alphabet of input symbols $\Sigma$
        \item a transition relation $\delta \subseteq (Q\times \Sigma) \times Q$
        \item an initial or start state called $s\in Q$
        \item a set of accepting states $F \subseteq Q$
    \end{enumerate}
    A BA accepts an infinite run iff at least one of its infinitely visited states is in $F$.
\end{definition}

\subsection{$\omega$-regular expressions}
\begin{definition}[Regular expression]\label{def:regex}
    Let $a$ be a symbol in $\Sigma$, $A$ be a regular expression, and $\epsilon$ be the empty expression. We define regular expressions using the following grammar:
    \begin{align*}
        A = \emptyset \mid \epsilon \mid a \mid AA \mid A + A \mid A^*
    \end{align*}
\end{definition}
\begin{definition}[Semantics of regular expressions]
    Let $\L(A)$ denote the set of propositional formulas accepted by regular expression $A$, also called the language accepted by $A$. We define $\L(A)$, inductively as:
    \begin{align*}
        \L(\emptyset) & = \emptyset \\
        \L(\epsilon) & = \{\epsilon\} \tag{$\epsilon$ denotes the empty string}\\
        \L(a) & = \{a\} \\
        \L(A_1A_2) & = \{s_1s_2 \mid s_1 \in L(A_1) \text{ and } s_2 \in L(A_2)\} \\
        \L(A_1 + A_2) & = L(A_1) \cup L(A_2) \\
        \L^{(0)}(A) & = \{\epsilon\} \\
        \L^{(i + 1)}(A) & = \{s_1s_2 \mid s_1 \in L(A) \text{ and } s_2 \in L^{(i)}(A)\} \\\
        \L(A^*) & = \bigcup_{i \ge 0} L^{(i)}(A)
    \end{align*}
\end{definition}

\begin{remark}
    For the purpose of our tool, we define $\Sigma$ as the set of propositional logic formulas:
    \[
    a \in \Sigma = p \mid \top \mid \bot \mid \neg \phi \mid \phi \land \psi \mid \phi \lor \psi \mid \phi \limplies \psi
    \]
\end{remark}

\begin{definition}[$\omega$-regular expressions]\label{def:omega-regex}
    Our definition \ref{def:regex} of regular expressions concerns only finite-length strings. However, since \ltl\ formulas reason about events that happen over an infinite-length timeline, we need to model them using infinite regular expressions (i.e., $\omega$-regular expressions), which we define via the following grammar:
    \begin{align*}
        B = A^{\omega} \mid AB \mid B + B
    \end{align*}
\end{definition}
\begin{definition}[Semantics of $\omega$-regular expressions]\label{def:omega-semantics}
    Let $\Sigma^{\omega}$ denote the set of infinite-length strings over fixed alphabet $\Sigma$. Let $\L_{\omega}(B)$ denote the $\omega$-language accepted by $\omega$-regular expression $B$. Then $\L_{\omega}(B)$ is inductively defined as
    \begin{align*}
        \L_{\omega}(A^{\omega}) & = \{s_1s_2s_3\cdots \mid s_i \in \L(A) \text{ and } i \ge 1\} \tag{$\epsilon
         \not\in \L(A)$} \\
        \L_{\omega}(AB) & = \{s_1s_2 \mid s_1 \in \L(A) \text{ and } s_2 \in \L_{\omega}(B)\} \\
        \L_{\omega}(B_1 + B_2) &= \L_{\omega}(B_1) \cup \L_{\omega}(B_2)
    \end{align*}
\end{definition}

@@

\subsection{$\omega$-regex to timeline} \label{regex2timeline}
Our tool uses Graphviz~\cite{Ellson2001GraphvizO} to achieve the timeline visualization step. By construction of our algorithm, every $\omega$-regular expression is in the form of
\[
    A_1A_2^{\omega} + A_3A_4^{\omega} + \cdots + A_{2n-1}A_{2n}^{\omega}
\]
where $A_i$ are regular expressions, $A_{2i-1}$ could be $\epsilon$, and $\epsilon \not\in \L(A_{2i})$. At a high level, for each of regular expression $A_i$, we visualize it as a set of accepted input. We view each union operator as parallel timelines, and each $A_{2i-1}$ is concatenated with $A_{2i}$ repeated infinitely many times, as represented in the grey box. \cref{fig:timeline} presents a generic timeline resulting from this form of construction.
\begin{figure}[h!]
    \centering
    \begin{tikzpicture}
    \node[draw, thick, fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a1) {$A_1$};
    \node[shape=rectangle, fill=gray!30!white, style={minimum width=1.8cm, minimum height=1.3cm}] (a2g) [right= 0.85cm of a1] {};
    \node[fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a2) [right= 1cm of a1] {$A_2$};
    \node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a3) [below= 0.4cm of a1] {$A_3$};
    \node[shape=rectangle, fill=gray!30!white, style={minimum width=1.8cm, minimum height=1.3cm}] (a4g) [right= 0.85cm of a3] {};
    \node[fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a4) [right= 1cm of a3] {$A_4$};
    \node[draw, thick, fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a5) [below= 0.4cm of a3] {...};
    \node[draw, thick, fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a7) [below= 0.4cm of a5] {\!\!\!$A_{2n-1}$\!\!\!};
    \node[shape=rectangle, fill=gray!30!white, style={minimum width=1.8cm, minimum height=1.3cm}] (a8g) [right= 0.85cm of a7] {};
    \node[fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (a8) [right= 1cm of a7] {\!\!\!$A_{2n}$\!\!\!};
    \node (m) at ($(a1)!0.5!(a7)$) {};
    \node[draw, thick, fill=white, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (s) [left= 1.5cm of m] {\!\!\!\!start\!\!\!\!};
    \draw[thick] (s) edge[-latex] (a1) (s) edge[-latex] (a3) (s) edge[-latex] (a5) (s) edge[-latex] (a7);
    \draw[thick] (a1) edge[-latex] (a2) (a3) edge[-latex] (a4) (a7) edge[-latex] (a8);
    \end{tikzpicture}
%    \includegraphics[scale=0.5]{img/timeline.png}
    \caption{Generic timeline construction from $A_1A_2^{\omega} + A_3A_4^{\omega} + \cdots + A_{2n-1}A_{2n}^{\omega}$}
    \label{fig:timeline}
\end{figure}

@@

\begin{lemma}
    For any NFA $G$ with start state $s$ and (exactly one) final state $f$, \textbf{nfa2regex\_firstvisit}$(G, s, f)$ outputs the regular expression of all paths from $s$ to $f$, visiting $f$ for the first time.
\end{lemma}
\begin{proof}
    Let $G'$ be $G$ with all out-edge (including loops) from $f$ deleted. We claim that $\L(G')$ is equal to the language $\L'$ of all paths from $s$ reaching $f$ for the first time.

    First we show $\L(G') \subseteq \L'$, i.e., every path accepted by $G'$ starts in $s$ and ends by reaching $f$ for the first time. Since $s$ and $f$ are the initial and final states of $G'$ respectively, clearly every path accepted by $G$ must start in $s$ and end in $f$. Further, as $f$ has no out-edges in $G$, any accepted path must end immediately once it reaches $f$. Thus it must end once it reaches $f$ for the first time.

    Also, we show $\L' \subseteq \L(G')$, i.e., every path starting in $s$ and ending by $f$ reaching $f$ for the first time is accepted by $G'$. If a path contains an out-edge of $f$, it cannot be in $\L'$ as the trace continues beyond the first time step at $f$. Thus, every trace in $\L'$ only uses edges in $G'$, and is thus accepted by $G$ iff it is accepted by $G'$.
\end{proof}

\begin{theorem}
    For any \Buchi automaton $G$ with start state $s$ and final states $F$, \textbf{ba2wregex}$(G)$ outputs the $\omega$-regular language accepted by $G$
\end{theorem}
\begin{proof}
    We show that
    \[
        \L = \bigcup_{f\in F} \textbf{nfa2regex}(G, s, f) \verb|^| \left(\textbf{nfa2regex\_firstvisit}(G, f, f)\right)^\omega
    \]
    and the $\omega$-language accepted by $G$, $\L_\omega(G)$ are identical.

    First, $\L\subseteq \L_\omega(G)$ since for every infinite run in $\L$, clearly by definition of $\L$, there is some $f\in F$ such that passes through $f$ infinitely often. The converse is true as well: for every infinite run $q_0, q_1, \dots$ accepted by $G$, then $q_0 = s$ and by definition of the accepting language, there is some accepting state $f$, passed through infinitely often, thus the run is in $\textbf{nfa2regex}(G, s, f)$ \verb|^| $\left(\textbf{nfa2regex\_firstvisit}(G, f, f)\right)^\omega$
\end{proof}
\subsection{Correctness of Rewrite Rules}
\begin{lemma}
    For all regular expressions $r_1, r_2, r$, each of the following rewrite rules preserve the regular or $\omega$-regular language represented by the expressions:
    \begin{alignat}{2}
        & r_1 + r_1r_2^* && \Longrightarrow r_1r_2^* \\
        & r + r && \Longrightarrow r \\
        & r_1 + r_2^*r_1 && \Longrightarrow r_2^*r_1 \\
        & (r^*)^{\omega} && \Longrightarrow r^{\omega} \\
        & (r_1r_2^*)r_2^{\omega} && \Longrightarrow r_1r_2^{\omega} \\
        & (r_1r_2)r_2^{\omega} && \Longrightarrow r_1r_2^{\omega} \\
        & r^*r^{\omega} && \Longrightarrow r^{\omega} \\
        & rr^{\omega} && \Longrightarrow r^{\omega}
    \end{alignat}
\end{lemma}

@@

\appendix
\section{Concrete Syntax of \tool} \label{sec:concrete-syntax}
Recall from~\ref{ltl-defn} that we define an \ltl\ formula as
\begin{align*}
    \varphi = & p \mid \neg \varphi \mid \varphi \land \psi \mid \varphi \lor \psi \mid \varphi \limplies \psi \mid \always \varphi \mid \eventually \varphi \mid \nextt \varphi \\ & \mid \varphi \stronguntil \psi \mid \varphi \weakrelease \psi
\end{align*}
where $p \in \AP$ and $\varphi$ and $\psi$ are LTL formulas. In our tool \tool, we use the following concrete input syntax to represent \ltl\ formulas in ASCII text.
\[
    \begin{array}{c r l l l@{\qquad} l}
         &      & \textit{abstract syntax}              & \textit{concrete syntax}                       & \textit{description} \\
    \varphi & =  & p                                     & \texttt{p}                                     & \text{atomic prop.} \\
         & \mid & \neg \varphi                             & \texttt{!} \varphi                                & \text{negation} \\
         & \mid & \varphi \land \psi                       & \varphi \texttt{ \& } \psi                        & \text{conjunction} \\
         & \mid & \varphi \lor \psi                        & \varphi \texttt{ | } \psi                         & \text{disjunction} \\
         & \mid & \varphi \limplies \psi                   & \varphi \texttt{ -> } \psi                        & \text{implication} \\
         & \mid & \always \varphi                          & \texttt{G} \varphi                                & \text{globally} \\
         & \mid & \eventually \varphi                      & \texttt{F} \varphi                                & \text{in the future} \\
         & \mid & \nextt \varphi                           & \texttt{X} \varphi                                & \text{next} \\
         & \mid & \varphi \stronguntil \psi                & \varphi \texttt{ U } \psi                         & \text{until} \\
         & \mid & \varphi \weakrelease \psi                & \varphi \texttt{ R } \psi                         & \text{release} \\
    \end{array}
\]
%\nocite{*}
\bibliographystyle{IEEEtran}
\bibliography{RelatedWork}
